Model for DES Type Ia Supernova Cosmology Analysis

### 2015-08-22 doc/model.tex
Note for U Penn folks

\begin{itemize}
\item All transients have measured counts at peak brightness $\mathit{ADU}$.
For transients in set {\it mis} without active spectroscopy:
\begin{multline}
p(\mathit{ADU}| z, \Omega_M, w, \theta_r, \alpha_{Ia},\sigma_{Ia}, \alpha_{\mathit{non-Ia}},\sigma_{\mathit{non-Ia}})=\\
\theta_r\mathcal{N}\left(\frac{\alpha_{Ia}}{d_L^2(z;\Omega_M, w)}, \frac{\ln{10}}{2.5}\sigma_{Ia}\right)+\left(1-\theta_r\right)
\mathcal{N}\left(\frac{\alpha_{\mathit{non-Ia}}}{d_L^2(z;\Omega_M, w)}, \frac{\ln{10}}{2.5}\sigma_{\mathit{non-Ia}}\right).
\end{multline}
The luminosity distance $d_L$ is for a flat universe.  
\item The transients of subset {\it obs} have spectroscopic observations of type $T_S$ and redshift
$z_S$, which are taken to be unambiguous.
\begin{itemize}
\item A direct association of $z_S=z$ is made.
\begin{equation}
p(z_S|z) = \delta(z_S-z)
\label{specz:eqn}
\end{equation}
\item For transients in set $T_S=1$:
\begin{equation}
p(\mathit{ADU}, T_S=1 | z, \Omega_M, w, \theta_r, \alpha_{Ia},\sigma_{Ia})=
\theta_r\mathcal{N}\left(\frac{\alpha_{Ia}}{d_L^2(z;\Omega_M, w)}, \frac{\ln{10}}{2.5}\sigma_{Ia}\right).
\end{equation}
\item For transients in set $T_S=0$:
\begin{multline}
p(\mathit{ADU}, T_S=0 | z, \Omega_M, w, \theta_r, \alpha_{\mathit{non-Ia}},\sigma_{\mathit{non-Ia}})= \\
\left(1-\theta_r\right)
\mathcal{N}\left(\frac{\alpha_{\mathit{non-Ia}}}{d_L^2(z;\Omega_M, w)}, \frac{\ln{10}}{2.5}\sigma_{\mathit{non-Ia}}\right).
\end{multline}
\end{itemize}
\item All transients have a host with redshift $z$ and have a second
galaxy at $z_N$ whose projection is close to the host, a potential source of ambiguity.
The pdf for the redshift of the second galaxy is approximately uniform in volume:
\begin{equation}
\frac{3 z_N^2}{z_{max}^3 - z_{min}^3}.
\end{equation}

@@

\item This simple example does not consider data $\theta_S$, $\tau$ or $Z$, i.e.
spectroscopic parameters, sample-selection effects, and calibration.
\end{itemize}

A prior is applied to the SN~Ia absolute magnitude
\begin{equation}
\alpha_{SNIa} \sim \ln\mathcal{N}\left(\ln\left(\alpha^0_{SNIa}\right),\frac{\ln{10}}{2.5}0.02\right),
\end{equation}
where $\alpha^0_{SNIa}$ is the true input luminosity.

Note some of the choices for the distributions are driven by computational
speed. 
Also, Stan cannot handle discrete parameters and
cannot except pdf's that support zero probability.

@@

The results that follow are based on the parameter
choices:
\begin{itemize}
\item Cosmological parameters $\Omega_M=0.28$, $w=-1$.
\item Type~Ia supernovae with intrinsic dispersion $\sigma_{SNIa}=0.1$.
\item Non-Ia supernovae 2 magnitudes fainter than SNe~Ia with intrinsic
dispersion $\sigma_{non-Ia}=1$.
\item $\theta_r=0.8$ for a constant relative fraction of SNe~Ia.
\item 500 transients uniformly distributed in the range $0.1<z<1$.
\item $z_{min}=0.05$ and $z_{max}=1.5$ is the range of possible true redshifts.
\item $P_{host}=0.98$ for the probability of correct host-galaxy assignment.
\item Accumulated spectra for 300, 400, and 500 transients.
\end{itemize}

The simulated data are analyzed using 
\citet{stan-software:2015}, a
a probabilistic programming language for
inference of Bayesian models with Hamiltonian Monte Carlo.  There are
four chains run, each with 50,000 samples then thinned by 50.  Half the links
are used for burnin, so ultimately 2000 links are used as draws from the posterior.
It took 2 days tor produce these results on my MacBook Air.  One chain, \#2
in the 50\% run,  took the bulk of the time; presumably a different integration step size
was determined during warmup.  As shown shortly, this chain samples
a broader range of parameter space.

\subsubsection{SN~Ia Only}
A classical analysis would only include confirmed SNe~Ia with a definitive
redshift. The realization of simulated data has 156 Type~Ia supernovae.
Results of the analysis of a pure SN~Ia sample with no
host misclassification are presented:  Figure~\ref{contour.100:fig}
shows the posteriors for  $\Omega_M$--$w$ and Table~\ref{ia_only:tab} the
credible equal-tailed intervals for $w$.  (Note that $w$ is the parameter
of most interest for the experiment.)
The distributions
look similar to the latest results from real data \citep{2014A&A...568A..22B} 
and so satisfy qualitative expectations.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[scale=0.55]{/Users/akim/project/abc/results/{contour..ia_only.200}.pdf} \newline
   \includegraphics[scale=0.55]{/Users/akim/project/abc/results/{contour.100}.pdf} 
   \includegraphics[scale=0.55]{/Users/akim/project/abc/results/{contour.200}.pdf}    \caption{Probability distribution function for $\Omega_M$--$w$ for: 156 SNe~Ia
   with spectra (top);  SNe~Ia and non-Ia with 50\% (left),
   and 100\% (right)  spectroscopic completeness.
   Solid lines represent the input cosmology.  Dashed lines in the
   histograms are the 0.16 and 0.86 quantiles. 
   \label{contour.100:fig}}
\end{figure}

\begin{table}
\centering
\begin{tabular}{|c|cc|}
\hline
\%-ile & $w$ range & $\Delta$\\ \hline
$0.68$ & $[-1.038, -0.648]$ & $0.389$ \\
$0.90$ & $[-1.204, -0.577]$ & $0.627$ \\
$0.95$ & $[-1.269, -0.556]$ & $0.713$ \\
\hline
\end{tabular}
\caption{Credible intervals of $w$ using the 156 SNe~Ia only with spectroscopic
observations.\label{ia_only:tab}}
\end{table}

\subsubsection{Full Sample: 50\% Spectroscopic Confirmation}
We now turn to the full sample of SNe~Ia and non-Ia's.  Recall that our simple
model treats the non-Ia's in the sample
as standard candles with a large (1 mag) intrinsic dispersion.

Results for the $\Omega_M$--$w$ posterior for the
case of 50\% spectroscopic completeness
are shown in Figure~\ref{contour.100:fig}.
As will be discussed shortly, the Stan run does not result in a Markov Chain Monte Carlo
that has
converged to a stationary solution.

A transient
without a spectrum has uncertain type and redshift: the redshift
of the associated host galaxy, and redshifts expected based on the
flux of the SN~Ia and non-Ia standard candles,
are viable but disjoint possibilities for the posterior of
the true underlying redshift.  To test the mixing of the chains,
the initial conditions 
for redshifts of transients without an active spectrum are
selected to be near local optima: the true redshift
for chains 0 and 2 and near the host redshift in chains 1 and 3.

Two objects, one Type~Ia and the other
non-Ia, are assigned incorrect redshifts (recall 50\% do not have
an active spectrum and the 98\% host misclassification rate).
Figure~\ref{missed_wrong.100:fig} shows draws
for the redshift parameter for these two transients.
The SN~Ia is placed at $z=0.95$ but is misassociated with a galaxy at
$z=1.24$.  The posterior shows a preference for the correct redshift, but sometimes
drops to lower redshifts as if it were a non-Ia. Although two of the chains start near the incorrect-host redshift, they do not return to that area of parameter space after burnin.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[scale=0.55]{/Users/akim/project/abc/results/{missed_wrong.100}.pdf} 
   \caption{Redshift draws for a Type~Ia and a non-Ia that were not observed
   spectroscopically and for whom the host-galaxies are misidentified.
   The draws of the four chains are appended in series.   \label{missed_wrong.100:fig}}
\end{figure}

The non-Ia is placed at $z=0.39$ and is misassociated with a galaxy at $z=0.48$.
Three of the four chains stay in the neighborhood of the incorrect-host redshift.
Chain 2, whose initial condition was at the true redshift, also accepts redshift solutions as
if were a Type~Ia. This
object is a source of bias as the region around the true redshift is never sampled.

Clearly Chain 2 does not mix with the others, so I compare some of its behavior with
Chains 0, 1, and 3 combined.
Figure~\ref{posterior.100:fig} shows the log-posteriors
of the two sets; the values are consistent so neither can be discounted as being trapped
in a local but not global optimum.
The $w$ draws from
Chains 0, 1, and 3, and those from Chain 2 are shown in Figure~\ref{posterior.100:fig},
and credible intervals in Table~\ref{split:tab}.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[scale=0.55]{/Users/akim/project/abc/results/{posterior.100}.pdf} 
   \includegraphics[scale=0.55]{/Users/akim/project/abc/results/{wsub.100}.pdf} 
   \caption{Log-posteriors and $w$ draws from the grouping of Chains 0, 1, and 3, 
and Chain 2. \label{posterior.100:fig}}
\end{figure}

\begin{table}[htbp] 
\centering
\begin{tabular}{|cc|cc|}
\hline
Chains & \%-ile & $w$ range & $\Delta$\\ \hline
013& $0.68$ & $[-1.029, -0.649]$ & $0.379$ \\
013& $0.90$ & $[-1.201, -0.579]$ & $0.623$ \\
013& $0.95$ & $[-1.290, -0.557]$ & $0.733$ \\ \hline
2& $0.68$ & $[-1.048, -0.673]$ & $0.375$ \\
2& $0.90$ & $[-1.197, -0.600]$ & $0.597$ \\
2& $0.95$ & $[-1.282, -0.575]$ & $0.707$ \\
\hline
\end{tabular}
\caption{Credible intervals of $w$ for the draws from the grouping of Chains 0, 1, and 3, 
and Chain 2.\label{split:tab}}
\end{table}

\subsubsection{Full Sample: Varying Spectroscopic Completeness}
The $\Omega_M$--$w$ posterior for the
case where all transients have a direct spectrum
is shown in Figure~\ref{contour.100:fig}.  The credible intervals
of $w$ for 50\%, 75\%, and 100\% spectroscopic completeness
are given in Table~\ref{compare:tab}.

\begin{table}[htbp] 
\centering
\begin{tabular}{|cc|cc|}
\hline
\# Spectra & \%-ile & $w$ range & $\Delta$\\ \hline
300& $0.68$ & $[-1.235, -0.943]$ & $0.292$ \\
300& $0.90$ & $[-1.330, -0.852]$ & $0.478$ \\
300& $0.95$ & $[-1.373, -0.806]$ & $0.568$ \\
400& $0.68$ & $[-1.237, -0.946]$ & $0.291$ \\
400& $0.90$ & $[-1.332, -0.858]$ & $0.474$ \\
400& $0.95$ & $[-1.379, -0.819]$ & $0.560$ \\
500& $0.68$ & $[-1.227, -0.945]$ & $0.282$ \\
500& $0.90$ & $[-1.324, -0.838]$ & $0.486$ \\
500& $0.95$ & $[-1.375, -0.785]$ & $0.590$ \\
\hline
\end{tabular}
\caption{Credible intervals of $w$  for the draws from the grouping of Chains 0, 1, and 3, 
and Chain 2.\label{compare:tab}}
\end{table}

\subsection{Proto-Conclusions}
\begin{itemize}
\item Host-galaxy misclassification and multiple populations can produce disjoint local maxima
in the posterior.  Sampling the space is a challenge.
\item For the simple model considered, statistical uncertainties dominate for the case
of 200 transients.
\end{itemize}

For the moment, the priority is to generate robust results rather than improve
the fidelity of the model.
\begin{itemize}
\item Run with $\sim 1000$ transients, which is more in line with what DES will get long-term
and will be less sensitive to realization fluctuations.
\item Save $\gg 2000$ links  of the chain to improve the accuracy of the posterior.
\item Achieve mixing of the chains within a reasonable compute time.  Possibilities
include reparametrization, optimization of the Stan code (I have already put non-trivial
effort into this), use of a different MCMC algorithm.
%\item Explore effect of galaxy-misclassification rate.  Its value is fairly uncertain at the moment.
\end{itemize}

>>>>

\begin{itemize}
\item All transients have measured counts at peak brightness $\mathit{ADU}$.
For transients in set {\it mis} without active spectroscopy:
\begin{multline}
p(\mathit{ADU}| z, \Omega_M, w, \theta_r, \alpha_{Ia},\sigma_{Ia}, \alpha_{\mathit{non-Ia}},\sigma_{\mathit{non-Ia}})=\\
\theta_r\mathcal{N}\left(\frac{\alpha_{Ia}}{d_L^2(z;\Omega_M, w)}, \frac{\ln{10}}{2.5}\sigma_{Ia}\right)+\left(1-\theta_r\right)
\mathcal{N}\left(\frac{\alpha_{\mathit{non-Ia}}}{d_L^2(z;\Omega_M, w)}, \frac{\ln{10}}{2.5}\sigma_{\mathit{non-Ia}}\right).
\end{multline}
The luminosity distance $d_L$ is for a flat universe.  
\item The transients of subset {\it obs} have spectroscopic observations of type $T_S$ and redshift
$z_S$, which are taken to be unambiguous.
\begin{itemize}
\item A direct association of $z_S=z$ is made.
\begin{equation}
p(z_S|z) = \delta(z_S-z)
\label{specz:eqn}
\end{equation}
\item For transients in set $T_S=1$:
\begin{equation}
p(\mathit{ADU}, T_S=1 | z, \Omega_M, w, \theta_r, \alpha_{Ia},\sigma_{Ia})=
\theta_r\mathcal{N}\left(\frac{\alpha_{Ia}}{d_L^2(z;\Omega_M, w)}, \frac{\ln{10}}{2.5}\sigma_{Ia}\right).
\end{equation}
\item For transients in set $T_S=0$:
\begin{multline}
p(\mathit{ADU}, T_S=0 | z, \Omega_M, w, \theta_r, \alpha_{\mathit{non-Ia}},\sigma_{\mathit{non-Ia}})= \\
\left(1-\theta_r\right)
\mathcal{N}\left(\frac{\alpha_{\mathit{non-Ia}}}{d_L^2(z;\Omega_M, w)}, \frac{\ln{10}}{2.5}\sigma_{\mathit{non-Ia}}\right).
\end{multline}
\end{itemize}
\item All transients have a host with redshift $z$ and have a second
galaxy at $z_N$ whose projection is close to the host, a potential source of ambiguity.
The pdf for the redshift of the second galaxy is approximately uniform in volume:
\begin{equation}
\frac{3 z_N^2}{z_{max}^3 - z_{min}^3}.
\end{equation}
The distribution of these galaxies is different from than the distribution of transient redshifts,
which can lead to bias.

@@

\item This simple example does not consider data $\theta_S$, $\tau$ or $Z$, i.e.
spectroscopic parameters, sample-selection effects, and calibration.
\item A prior is applied to the SN~Ia absolute magnitude
\begin{equation}
\alpha_{SNIa} \sim \ln\mathcal{N}\left(\ln\left(\alpha^0_{SNIa}\right),\frac{\ln{10}}{2.5}0.02\right),
\end{equation}
where $\alpha^0_{SNIa}$ is the true input luminosity.
\end{itemize}

I highlight the treatment of redshift in this model; the cosmological redshift
$z$ is set to the values of the
observed redshifts through $\delta$-functions.  This choice was made specifically to aid
in the convergence of the Monte Carlo: the discrete possibilities for redshift allow
for their marginalized likelihood to be used in the calculation of the posterior.
A more general model would replace the $\delta$-functions
with coming from a continuous PDF's.
Peculiar velocities, redshift measurement errors (particularly from
photometric redshifts) justify this generalization.
A transient
without a spectrum has uncertain type and redshift: the redshift
of the associated host galaxy, and redshifts expected based on the
flux of the SN~Ia and non-Ia standard candles,
are viable but disjoint possibilities for the posterior of
the true underlying redshift. My runs of this scenario produce chains
that do not mix so I resorted to the simplified treatment presented here.

@@

The results that follow are based on the parameter
choices:
\begin{itemize}
\item Cosmological parameters $\Omega_M=0.28$, $w=-1$.
\item Type~Ia supernovae with intrinsic dispersion $\sigma_{SNIa}=0.1$.
\item Non-Ia supernovae 2 magnitudes fainter than SNe~Ia with intrinsic
dispersion $\sigma_{non-Ia}=1$.
\item $\theta_r=0.8$ for a constant relative fraction of SNe~Ia.
\item 500 transients uniformly distributed in the range $0.1<z<1$.
\item $z_{min}=0.05$ and $z_{max}=1.5$ is the range of possible misassigned
host redshift.
\item $P_{host}=0.98$ for the probability of correct host-galaxy assignment.
\item Accumulated spectra for 200, 350, and 500 transients.
\end{itemize}

The simulated data are analyzed using 
\citet{stan-software:2015}, a
a probabilistic programming language for
inference of Bayesian models with Hamiltonian Monte Carlo.  There are
four chains run, each with 50,000 samples then thinned by 5.  Half the links
are used for burnin, so ultimately 20,000 links are used as draws from the posterior.

\subsubsection{SN~Ia Only}
A classical analysis would only include confirmed SNe~Ia with a definitive
redshift. The realization of simulated data has 415 Type~Ia supernovae.
Results of the analysis of a pure SN~Ia sample with no
host misclassification are presented:  Figure~\ref{contour.ia_only:fig}
shows the posteriors for  $\Omega_M$--$w$ and Table~\ref{ia_only:tab} the
credible equal-tailed intervals for $w$.  The true $w=-1$ is within
the 0.58 credible interval. (Note that $w$ is the parameter
of most interest for the experiment.)

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[scale=0.55]{/Users/akim/project/abc/results/{contour..ia_only.500}.pdf} \newline
\caption{Probability distribution function for $\Omega_M$--$w$ for only SNe~Ia
   all with spectra.
   Solid lines represent the input cosmology.  Dashed lines in the
   histograms are the 0.16 and 0.86 quantiles. 
   \label{contour.ia_only:fig}}
\end{figure}

\begin{table}
\centering
\begin{tabular}{|c|cc|}
\hline
\%-ile & $w$ range & $\Delta w$\\ \hline
$0.68$ & $[-1.254, -0.975]$ & $0.279$ \\
$0.90$ & $[-1.347, -0.889]$ & $0.458$ \\
$0.95$ & $[-1.396, -0.846]$ & $0.550$ \\
\hline
\end{tabular}
\caption{Credible intervals of $w$ using the SNe~Ia only with spectroscopic
observations.\label{ia_only:tab}}
\end{table}

\subsubsection{Full Sample: Varying Spectroscopic Completeness}
We now turn to the full sample of SNe~Ia and non-Ia's.  Recall that our simple
model treats the non-Ia's in the sample
as standard candles with a large (1 mag) intrinsic dispersion.

Results for the $\Omega_M$--$w$ posterior for the
case of 40\%, 70\%, and 100\% spectroscopic completeness
are shown in Figures~\ref{contour.200:fig}, \ref{contour.350:fig}, and
\ref{contour.500:fig} respectively.
The credible intervals
of $w$ for are given in Table~\ref{compare:tab}, with
the input $w=-1$ within the 0.65, 0.49, and 0.47  intervals
respectively.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[scale=0.40]{/Users/akim/project/abc/results/{contour.200}.pdf} 
   \caption{Probability distribution function for $\Omega_M$--$w$ for:
    SNe~Ia and non-Ia with 40\%  spectroscopic completeness.
   \label{contour.200:fig}}
\end{figure}

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[scale=0.40]{/Users/akim/project/abc/results/{contour.350}.pdf} 
   \caption{Probability distribution function for $\Omega_M$--$w$ for
    SNe~Ia and non-Ia with 70\%  spectroscopic completeness.
   \label{contour.350:fig}}
\end{figure}

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[scale=0.40]{/Users/akim/project/abc/results/{contour.500}.pdf} 
   %\includegraphics[scale=0.55]{/Users/akim/project/abc/results/{contour.200}.pdf}    
   \caption{Probability distribution function for $\Omega_M$--$w$ for
    SNe~Ia and non-Ia with 100\%  spectroscopic completeness.
   \label{contour.500:fig}}
\end{figure}

\begin{table}[htbp] 
\centering
\begin{tabular}{|cc|cc|}
\hline
\# Spectra & \%-ile & $w$ range & $\Delta$\\ \hline
200& $0.68$ & $[-1.284, -0.991]$ & $0.293$ \\
200& $0.90$ & $[-1.385, -0.900]$ & $0.485$ \\
200& $0.95$ & $[-1.435, -0.860]$ & $0.575$ \\
350& $0.68$ & $[-1.243, -0.952]$ & $0.291$ \\
350& $0.90$ & $[-1.339, -0.858]$ & $0.481$ \\
350& $0.95$ & $[-1.390, -0.817]$ & $0.573$ \\
500& $0.68$ & $[-1.236, -0.948]$ & $0.289$ \\
500& $0.90$ & $[-1.336, -0.861]$ & $0.475$ \\
500& $0.95$ & $[-1.385, -0.815]$ & $0.570$ \\
\hline
\end{tabular}
\caption{Credible intervals of $w$ given different levels of spectroscopy.  \label{compare:tab}}
\end{table}

As would be expected, the better the spectroscopic completeness, the less bias
and smaller uncertainties that are obtained.
